Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-06-21 analysis/paper_draft2.Rmd
Update paper_draft2.Rmd
updated equation 5 so that now it's using general notation
changed notation for initial prior in Eq 6; removed subscripts from mu and sigma; changed text to indicate Figure 5 assumes b_curr = b_hist; adjusted formatting of Table 1

\begin{equation}  
\pi_0(b) \sim Normal(\mu_0 = 0, \sigma_0 = 2)
\label{eq:initialprior}
\end{equation}

For this initial prior, we let $\mu_0 = 0$, because the law assigns the burden of proof in a _Batson_ challenge to the party bringing the challenge. Thus, if the challenging party produces no relevant evidence of illegal bias, the law requires a trial judge to reject the challenge as unproven. This is tantamount to treating zero as the most-likely value of the bias parameter, absent any data. Moreover, we take the law to imply that, absent any data, one must assume that higher degrees of illegal bias are less likely than lower degrees of bias. For this reason, we use a normal (Gaussian) distribution with $\sigma_0 = 2$. This prior accords with @Kadane2021 [p. 51], who suggests a prior for estimating _Batson_ strike bias that is at least unimodal, symmetric, and not dependent on the data.

@@

\begin{equation}
L(b|\delta) \propto L(b|\delta)(L(b|\delta_0))^{\alpha_0}\exp(-\frac{b^2}{8})
\label{eq:model4}
\end{equation} where

@@

Bias detection will also be sensitive to the choice of the standard deviation of the initial prior. The results above depend in part on our choice of a conservative initial prior for the bias parameter (Equation \@ref(eq:initialprior)). Given little historical strike data for an attorney and a low number of strikes per trial, this initial prior will dominate. As a result, the model we use will only detect severe bias. This sensitivity, however, reduces after including the historical data, especially when $\alpha$ is close to 1.

@@

Figure \@ref(fig:figsdprior) depicts how assigning different values for the initial prior’s standard deviation ($\sigma_0 = \{1,2,100\}$) affects the bias detection rate at the 95\% credible interval for three $\alpha$ values ($\alpha = \{0.1,0.5,1\}$) under scenarios where each attorney had 15 strikes and where $b_{curr} = b_{hist}$. In our simulations, we set $\sigma_0 = 2$ for our initial prior. If we instead select $\sigma_0 = 1$, that matters most when $\alpha = 0.1$, i.e., when we borrow less information from the historical data. As $\alpha$ increases, the initial prior's standard deviation matters less. Moreover, when we increased the initial prior's standard deviation to make it much less informative ($\sigma = 100$), that had little impact on bias detection rates, regardless of the choice of $\alpha$.

@@

Finally, Figure \@ref(fig:figcmptbly) summarizes our simulation results for bias detection at the 95% credible interval. These results indicate that our model can accurately detect strong bias if present in both the current and historical trials. Given high compatibility, increasing $\alpha$ improves bias detection, as does increasing the number of historical trials and more strikes per trial. Under the low number of strikes per trial (n = 6), bias detection requires strong bias in the current trial, a high $\alpha$, and high compatibility.

@@

knitr::kable(df, caption ="Hypothetical Strike Tally", format = "latex")

>>>>

\begin{equation}  
\pi_0(b) \sim Normal(\mu = 0, \sigma = 2)
\label{eq:initialprior}
\end{equation}

For this initial prior, we let $\mu = 0$, because the law assigns the burden of proof in a _Batson_ challenge to the party bringing the challenge. Thus, if the challenging party produces no relevant evidence of illegal bias, the law requires a trial judge to reject the challenge as unproven. This is tantamount to treating zero as the most-likely value of the bias parameter, absent any data. Moreover, we take the law to imply that, absent any data, one must assume that higher degrees of illegal bias are less likely than lower degrees of bias. For this reason, we use a normal (Gaussian) distribution with $\sigma = 2$. This prior accords with @Kadane2021 [p. 51], who suggests a prior for estimating _Batson_ strike bias that is at least unimodal, symmetric, and not dependent on the data.

@@

\begin{equation}
  L(b|\delta) \propto L(b|\delta)(L(b|\delta_0))^{\alpha_0}\exp(-\frac{(b-\mu)^2}{2\sigma^2})
  \label{eq:model4}
\end{equation} where

@@

Bias detection will also be sensitive to the choice of the standard deviation of the initial prior ($\sigma$). The results above depend in part on our choice of a conservative initial prior for the bias parameter (Equation \@ref(eq:initialprior)). Given little historical strike data for an attorney and a low number of strikes per trial, this initial prior will dominate. As a result, the model we use will only detect severe bias. This sensitivity, however, reduces after including the historical data, especially when $\alpha$ is close to 1.

@@

Figure \@ref(fig:figsdprior) depicts how assigning different values for the initial prior’s standard deviation ($\sigma = \{1,2,100\}$) affects the bias detection rate at the 95\% credible interval for three $\alpha$ values ($\alpha = \{0.1,0.5,1\}$) under scenarios where each attorney had 15 strikes and where $b_{curr} = b_{hist}$. In our simulations, we set $\sigma = 2$ for our initial prior. If we instead select $\sigma = 1$, that matters most when $\alpha = 0.1$, i.e., when we borrow less information from the historical data. As $\alpha$ increases, the initial prior's standard deviation matters less. Moreover, when we increased the initial prior's standard deviation to make it much less informative ($\sigma = 100$), that had little impact on bias detection rates, regardless of the choice of $\alpha$.

@@

Finally, Figure \@ref(fig:figcmptbly) summarizes our simulation results for bias detection at the 95% credible interval, given perfect compatibility ($b_{curr} = b_{hist}$). These results indicate that our model can accurately detect strong bias if present in both the current and historical trials. Given high compatibility, increasing $\alpha$ improves bias detection, as does increasing the number of historical trials and more strikes per trial. Under the low number of strikes per trial (n = 6), bias detection requires strong bias in the current trial and a high $\alpha$.

@@

knitr::kable(df, caption ="Hypothetical Strike Tally", format = "latex", booktabs = TRUE)
